\section{Functions}

Functions map a value from an input set - the \textit{domain} - to an output set - the \textit{co-domain}. Every output of the function is contained in a subset of the co-domain, called the \textit{range}.
Functions are \textit{pure}, meaning
\begin{itemize}
  \item The function returns identical values for identical inputs (i.e. there is no variation due to local variables).
  \item The function has no side-effects -- this is guaranteed due to data immutability.
\end{itemize}

In Haskell, occurences of functions are matched against patterns and expanded into their right-hand side.

\subsection{Function Definition}
Functions are defined by providing its name, a list of arguments, and setting it equal to an expression.

\begin{center}
  \texttt{\textcolor{blue}{name} arg1 arg2 \ldots argn = <expr>}
\end{center}

The arguments may be set to constants, or may be given a name to accept a variable value.

The type of a function is specified by an arrow (\texttt{->}) seperated list of its argument types and its return type:

\begin{center}
  \texttt{\blue{func} :: \red{type$_1$} -> \red{type$_2$} -> \ldots -> \red{type$_n$} -> \red{type$_\text{return}$}}
\end{center}

For an example, take a function which returns the sum of the elements of a list:

\begin{lstlisting}[language=haskell]
  sum :: [a] -> Int -- Takes a list of an arbitrary type and returns an integer
  sum [] = 0 -- Define the sum of an empty list to be zero (base case)
  sum (h:t) = h + sum t -- Define the sum of a list to be the head plus the sum of the tail
\end{lstlisting}

\subsubsection{Infix Functions}

A good example of infix functions are operators such as \texttt{+}. Functions which take two arguments may be written between the arguments instead.

For example, say we had \texttt{\blue{add} a b = a + b}. Then \texttt{\blue{add} 5 7} and \texttt{5 \;\textasciigrave\blue{add}\textasciigrave\; 7} are equivalent.

To reference the function defined by an operator i.e. \texttt{+}, surround it with parenthesis i.e. \texttt{(+)}.

The associativity and precedence of infix operators may be changed via \texttt{\blue{infixr}/\blue{infixl} <prec> <function name>} where $0 < \texttt{prec} \le 9$.

\subsection{Pattern Matching}

Parameters of functions may be matched against a pattern. Note that order matters: more specific patterns should be defined first, with general cases last.
\begin{itemize}
  \item Accept any value by using a symbol e.g. \texttt{f x = ...}.
  \item Accept a certain value e.g. \texttt{f 0 = ...}, \texttt{g [] = ...}
  \item List splicing using \texttt{(x:xs)} where \texttt{x} is the head, \texttt{xs} is the tail.
  \item Tuple unpacking e.g. \texttt{(Int,Int)} could be unpacked using \texttt{(x,y)}.
  \item Accept and extract a custom types. E.g., say we had \texttt{\blue{type} Pos = (Int, Int)}, we would define \texttt{getX (x, y) = x}.
  \item Accept and extract a custom datatype. E.g., say we had \texttt{\blue{data} Num = Zero | Succ Num}. We could then define \texttt{f Zero = ...} and \texttt{f (Succ n) = ...}.
\end{itemize}

\subsubsection{Top-Level}
Patterns may be defined in the topmost level.
\begin{lstlisting}[language=haskell]
f :: ...
f <pattern1> = <expr1>
f <pattern2> = <expr2>
...
\end{lstlisting}

\subsubsection{Guards}
Guards resemble piecewise functions.
\begin{lstlisting}[language=haskell]
f :: ...
f
  | <pattern1> = <expr1>
  | <pattern1> = <expr1>
  ...
  | otherwise = <expr_default> -- Catch-all
\end{lstlisting}

\subsubsection{Case Of}
\begin{lstlisting}[language=haskell]
case <expr> of
  <pattern1> -> <expr1>
  <pattern2> -> <expr2>
  ...
\end{lstlisting}

\subsubsection{If-Else}
Syntactic sugar, if-else expressions resemble conditional statements in other languages.
\begin{lstlisting}[language=haskell]
if <expr> then <ifTrue> else <ifFalse>
\end{lstlisting}
Elseif conditions may be constructed by chaining multiple if-else expressions after the else keyword.

Using a language extension, multiple conditions may be listed without chaining if-else expressions.
\begin{lstlisting}[language=haskell]
{-# LANGUAGE MultiWayIf #-}
if
  | <cond1> -> <expr1>
  | <cond2> -> <expr2>
  ...
  | otherwise -> <expr_default>
\end{lstlisting}

\subsection{Function Application}

A function is applied (called) to some arguments as follows:

\begin{center}
  \texttt{\blue{name} arg1 arg2 \ldots argn}
\end{center}

For example, consider a function \texttt{in\_range} which implements $x \in [min, max)$.
\begin{lstlisting}[language=haskell]
-- Function definition
in_range :: Int -> Int -> int -> Bool
in_range x min max = x >= min && x < max

-- Function application
in_range 3 0 5 = True
in_range 5 0 5 = False
\end{lstlisting}

\subsection{Recursion}
Recursion is the process of a function calling itself. Recursion requires a \textit{base case} to stop the function recursing indefinitely.

There are many ways to implement recursion, which will be demonstrated using the factorial function, defined as
\[n! = n \cdot (n - 1) \cdot \ldots \cdot 1 = \prod_{k = 1}^{n} k\]

\subsubsection{Defined Base Case}
We can hard-code the case where the function is called with the base case:

\begin{lstlisting}[language=haskell]
fac :: Int -> Int
fac 0 = 1
fac 1 = 1
fac n = n * fib (n-1) 
\end{lstlisting}

\subsubsection{If-Else Expression}
We can use the if-else expression:
\begin{lstlisting}[language=haskell]
fac :: Int -> Int
fac n = if n <= 1 then 1 else n * fac (n-1)
\end{lstlisting}

\subsubsection{Guards}
Guards are similar to piece-wise functions.
\begin{lstlisting}[language=haskell]
fac :: Int -> Int
fac n =
  | n <= 1    = 1
  | otherwise = n * fac (n-1)
\end{lstlisting}

\subsubsection{Accumulators}
In this example, we define an auxiliary function \texttt{\blue{aux}} inside \texttt{\blue{fac}} to calculate the the factorial

\begin{lstlisting}[language=haskell]
  fac n = aux n 1
    where
      aux n acc
        | n <= 1    = acc
        | otherwise = aux (n-1) (n*acc)
\end{lstlisting}

This is called \textit{tail recursion}. This is because the final result of \texttt{\blue{aux}} is the result we want, meaning that it is much more memory efficient. A good compiler could even unwind this into a non-recursive imperative approach using a loop. (For more insight, see \url{https://www.youtube.com/watch?v=_JtPhF8MshA}.)

Normal recursion (using an above definition of \texttt{\blue{fac}}):
\begin{verbatim}
fac 4
= 4 * fac 3
= 4 * (3 * fac 2)
= 4 * (3 * (2 * fac 1))
= 4 * (3 * (2 * 1))
= 4 * (3 * 2)
= 4 * 6
= 24
\end{verbatim}

Tail recursion (using the definition in this sub-section):
\begin{verbatim}
fac 4
= aux 3 4
= aux 2 12
= aux 1 24
= 24
\end{verbatim}

\subsection{Lambdas}
Syntax:
\begin{center}
  \texttt{(\blue{\symbol{92}}<args> -> <expr>)}
\end{center}

Some examples:
\begin{itemize}
  \item \texttt{(\symbol{92}x -> x+1) 2} returns \texttt{3}.
  \item \texttt{(\symbol{92}x y z -> x+y+z) 1 2 3} returns \texttt{6}.
\end{itemize}
Lambdas may be bound to names.

\subsection{Higher Order Functions}
Higher order functions are functions that take other functions as arguments.

For example, a function which takes another function and applies it to an argument:
\begin{lstlisting}[language=haskell]
  app :: (a -> b) -> a -> b
  app f x = f x
\end{lstlisting}
A synonym of such a function is the dollar (\texttt{\$}) operator: \texttt{(\$) :: (a -> b) -> a -> b}.

\subsubsection{Useful Higher Order Functions}
\begin{itemize}
  \item \texttt{\blue{map} :: (a -> b) -> [a] -> [b]} applies a function to every element on an array.
  \[L' = \{f(x) : x \in L\}\]
  
  Example: \texttt{\blue{map} (\symbol{92}x -> x\string^2) [1,2,3]} returns \texttt{[1,4,9]}.
  \item \texttt{\blue{filter} :: (a -> \red{Bool}) -> [a] -> [a]} filters the list on a predicate.
  \[L' = \{x \in L : P(x)\}\]
  
  Example: \texttt{\blue{filter} (\symbol{92}x -> mod x 2 == 0) [1,2,3,4,5]} returns \texttt{[2,4]}.
  \item \texttt{\blue{fold} :: (a -> b -> b) -> b -> [a] -> b} processes a list with some function to produce a single value, starting at a given value.
  
  In Haskell, \texttt{\blue{fold}} doesn't exist, but rather \texttt{\blue{foldr}} and \texttt{\blue{foldl}} wherein the position of the base value is different

  \begin{lstlisting}[language=haskell]
foldr f a [] = a
foldr f a (x:xs) = f x (foldr f a xs)
-- foldr f a [x1, x2, ..., xn] = (x1 `f` (x2 `f` (... (xn `f` a))))

foldl f a = []
foldl f a (x:xs) = f (foldl f a xs) x
-- foldl f a [x1, x2, ..., xn] = (...((a `f` x1) `f` x2) `f` xn)
  \end{lstlisting}

  See \texttt{code/folding.hs}

  Example: \texttt{\blue{foldr} (+) 0 [1,2,3,4,5]} returns \texttt{15}.
\end{itemize}

\subsection{Currying}
The principle behind currying is that given
\begin{center}
  \texttt{\blue{f} :: a -> b -> c -> d}
\end{center}
We could re-write this as
\begin{center}
  \texttt{\blue{f'} :: a -> (b -> (c -> d))}
\end{center}

For example, one could define a function \texttt{\blue{add}} in multiple ways:
\begin{lstlisting}[language=haskell]
  add x y = x + y
  add x = (\y -> x + y)
  add = (\x -> (\y -> x + y))
\end{lstlisting}

\subsubsection{Currying}
\begin{lstlisting}[language=haskell]
curry :: ((a,b -> c) -> a -> b -> c)
curry f x y = f (x,y)
\end{lstlisting}

\subsubsection{Uncurrying}
\begin{lstlisting}[language=haskell]
uncurry :: (a -> b -> c) -> (a,b) -> c
uncurry f (x,y) = f x y
\end{lstlisting}

\subsubsection{Partial Function Application}
Using the last definition of \texttt{\blue{add}}, consider the result of \texttt{\blue{add} 1}. This would be a new function: \texttt{\blue{add} 1 :: \red{Int} -> \red{Int}}. This is known as a \textit{section}.

A good example would be using \texttt{\blue{map}}.
\begin{center}
  \texttt{\blue{doubleList} = \blue{map} (\symbol{92}x -> x * 2)}
\end{center}

\subsection{Function Composition}
Function composition is a way to combine functions. For this, we use the dot (\texttt{.}) operator.
\begin{center}
  \texttt{(.) :: (b -> c) -> (a -> b) -> (a -> c)}
\end{center}
Then, \texttt{(f.g)} is equivalent to \texttt{(\blue{\symbol{92}}x -> f (g x))}.

For example, all three definitions of \texttt{\blue{descSort}} are equivalent:
\begin{lstlisting}[language=haskell]
  descSort = reverse . sort
  descSort = \x -> reverse (sort x)
  descSort x = reverse $ sort x
\end{lstlisting}
